Examining equity in accessibility to bike share: a balanced floating catchment area approach

### 2021-08-23 Bike-share-spatial-equity/Bike-share-spatial-equity.tex
Major revisions resubmitted June 2021

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Examining equity in accessibility to bike share - a balanced floating catchment area approach},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Examining equity in accessibility to bike share - a balanced floating
catchment area approach}
    \author[McMaster University]{Elise Desjardins\corref{1}}
   \ead{desjae@mcmaster.ca} 
    \author[University of Toronto Scarborough]{Christopher D. Higgins}
   \ead{cd.higgins@utoronto.ca} 
    \author[McMaster University]{Antonio Paez}
   \ead{paezha@mcmaster.ca} 
      \address[McMaster University]{School of Earth, Environment \& Society, McMaster University, 1280 Main
Street West, Hamilton, ON L8S4L8}
    \address[University of Toronto Scarborough]{Department of Geography \& Planning, University of Toronto Scarborough,
1265 Military Trail, Toronto, ON M1C1A4}
      \cortext[1]{Corresponding Author}

@@

Many PBSPs in North America now offer specific programs to address
equity that primarily focus on removing cost barriers and increasing
access for groups that are under-represented among existing users
(McNeil et al., 2019). The locations of docking stations is a common
consideration for increasing equity (Howland et al., 2017). Hamilton
Bike Share (HBS), located in Hamilton, Ontario, was the only Canadian
PBSP included in a North American scan of bike share equity programs
(see McNeil et al., 2019). HBS was launched in 2015 and currently has
over 900 operational bicycles and 130 docking stations. An equity
program, Everyone Rides Initiative, was implemented in 2018 which
expanded the program by introducing twelve ``equity'' stations which
expanded HBS to more disadvantaged neighbourhoods in the core service
area.

This paper examines accessibility to Hamilton Bike Share using a
balanced floating catchment area (BFCA) approach. We also conduct a
comparative analysis to the conventional two-step floating catchment
area approach to highlight the benefits of this method which, to our
knowledge, has not yet been used in the cycling literature. The paper
also assesses the contribution of HBS's equity stations to reducing
inequities in accessibility for different groups according to median
total household income, and provides policy recommendations to further
improve equity.

@@

\hypertarget{literature-review}{%
\section{2.1. Literature Review}\label{literature-review}}

\hypertarget{public-bicycle-share-programs}{%
\subsection{1.1. Public Bicycle Share
Programs}\label{public-bicycle-share-programs}}

@@

\hypertarget{equity-of-pbsps}{%
\subsection{1.2. Equity of PBSPs}\label{equity-of-pbsps}}

@@

On the whole, existing studies highlight the need for PBSPs to be more
accessible for diverse populations in order to increase use beyond the
``typical'' users. This has been the focus of recent research (see,
among others, Auchincloss et al., 2020; Hull Grasso et al., 2020;
MacArthur et al., 2020). Offering more people the option of using
sustainable and active transportation, particularly those who have lower
socioeconomic status and might benefit the most, is a worthy policy goal
for cities with PBSPs. However, exploring transportation equity by
investigating where bicycle share stations are located, often using
neighborhoods or census tracts as the geographical unit of analysis, can
ignore or miss the benefits that may be derived from adjacent zones.
Meaning that, stations may be lacking in certain neighborhoods but there
may be stations accessible within a reasonable walking time. This is
where geographical accessibility becomes an important consideration.

@@

The location and size of docking stations are two factors that are
relevant to accessibility. Since the time or distance needed to reach a
docking station decreases the potential of accessing the program, the
location of stations matters. Kabra et al.~(2020) found that the
majority of bike share usage in Paris comes from areas within 300 m of
stations, which amounts to 2-4 minutes walking by an adult who does not
have a disability. Similar to other public amenities affected by
crowding, the utility of stations is also limited by the \emph{maximum}
number of bicycles that they can hold (e.g., their size). Accessibility
analyses for PBSPs constitute a positive and evaluation-based approach
that also has the potential to inform equity efforts. For instance, Wang
and Lindsey (2019b) investigated whether new or relocated bicycle share
stations increased accessibility and use, which offered important
insights to improve the performance of the program.

Several approaches have been commonly used for measuring place-based
accessibility, including cumulative opportunities, gravity, and
utility-based measures (Handy and Niemeier, 1997). The gravity-based
approach involves weighting destination opportunities, such as the
quantity of bicycle share stations, by the time required to reach them
from an origin using an impedance function (Handy and Niemeier, 1997;
Kwan, 1998). While such measures are suitable for capturing the
potential for reaching destinations from a given location, they do not
take demand or congestion effects into account which is an important
consideration when calculating accessibility for amenities such as bike
share stations.

@@

When measuring accessibility, researchers have also taken different
approaches when it comes to the aggregation of data, either by using the
individual or household as the smallest unit of analysis or larger
spatial zones. Previous research on bicycle share equity has typically
used a meso- or macro-level approach with aggregated data from entire
neighborhoods or census tracts (Babagoli et al., 2019; Mooney et al.,
2019; Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), although there
are recent exceptions (Chen et al., 2019; Chen and Li, 2021). This is
also true for studies examining correlates of bicycle share demand (J.
Wang and Lindsey, 2019b). Handy and Niemeier (1997) note that using
disaggregated data in accessibility analyses provides a more accurate
estimate for individuals. Chen et al.~(2019, p. 530) are in favour of
using disaggregated data, which they did in their recent analysis of
Tampa's PBSP, because they note ``the use of aggregated data might
hinder our understanding of the equity impacts since individual
disparities are absorbed after aggregation''.

@@

Our analysis builds upon a previous study (Hosford and Winters, 2018),
which found that areas in Hamilton with less advantage are better served
by the city's PBSP compared to other Canadian cities (i.e., Toronto,
Vancouver, Montreal, and Ottawa-Gatineau) where areas that are less
deprived have greater access. Hosford and Winters (2018, pp.~)
acknowledge that ``Hamilton stands out in that the lower income
neighborhoods are located near the city center and wealthier
neighborhoods are in the surrounding suburban areas''. Therefore, the
core service area for the PBSP in Hamilton by default covers more of the
less advantaged areas in the city. However, there is also a great deal
of variation in income in the city center because of the local
university and increasing gentrification. Hosford and Winters (2018)
took a macro-level approach in their analysis by using dissemination
areas \emph{across} the city as the unit of analysis. They did not focus
specifically on the core service area and did not differentiate between
conventional and equity stations.

Therefore, this paper contributes to the cycling literature by expanding
the analysis of Hosford and Winters (2018) to assess equity in
accessibility to HBS and explore the contribution of the ``equity''
stations added to Hamilton's PBSP.

@@

The case study of this paper is the city of Hamilton, Ontario, Canada.
Before June 2020, the program was known as Social Bicycles or SoBi
Hamilton, but is now called Hamilton Bike Share (HBS). The core service
area spans 40 sq.km of the city, although it was planned to be 20 sq.km
(Hamilton, 2015a), and roughly 138,000 people are within 30 minutes
walking of a station {[}see Figure
\ref{fig:hamilton-and-sobi-service-area}{]}. This represents roughly one
fifth of the total population of the Hamilton Census Metropolitan Area
according to the 2016 Canadian Census. The City of Hamilton undertook a
large public engagement campaign to validate the locations of stations
that had been selected and to crowdsource potential locations for
additional locations (Hamilton, 2014). Most of the suggested locations
were in the east end of the core service area that lack stations or in
neighborhoods not serviced by the PBSP. The program was enthusiastically
welcomed in the city in 2015 - within three weeks of launching, 10,000
trips had been made (Hamilton, 2015b), however inadequate coverage given
the size of Hamilton Bike Share's service area was identified as a
problem early on, and transportation planners noted that the small size
(i.e., supply of bicycle racks) and low quantity of stations would lead
to challenges in balancing the system (Hamilton, 2015a).

@@

Hamilton Bike Share's equity program, Everyone Rides Initiative (ERI),
was launched in 2018 with the objective of reducing barriers that may
prevent individuals from accessing the program. Additional bicycles and
twelve ``equity'' stations were added to the core service area. The
equity program also offers subsidized memberships to individuals who
identify as low income, and complements this service with cycle skills
education. A comparable program can be found in Philadelphia (see Caspi
and Noland, 2019).

@@

\caption{The spatial distribution of bicycle share stations in Hamilton, Ontario. Conventional stations are in blue and equity (ERI) stations are in orange. The service area of Hamilton Bike Share is outlined in black and the city's downtown core is outlined in dark green. Hamilton Census Metropolitan Area is shown in grey.}\label{fig:sobi-stations-in-hamilton}
\end{figure}

@@

Since walking trips to a bicycle share station likely happen at a level
lower than even the smallest census geography, this requires a more
disaggregated approach than the use of census geographies. For this
reason, we implemented our analysis parting from micro population zones
to better reflect the friction of walking to a docking station, which is
an important component of a bike share trip (Chen et al., 2019).

To obtain population at sub-census geography levels, we used
pycnophylactic interpolation (Tobler, 1979). We obtained population data
from the 2016 Census of Canada for dissemination areas (DA), which is
the smallest publicly available census geography in Canada. These zonal
values of the population were interpolated to smaller polygons that are
50-by-50 m in size. Pycnophylactic interpolation involves smoothing out
the population from each DA while preserving total volume {[}see Figure
\ref{fig:da-population}{]}. When interpolating the population at this
high level of resolution, it is important to ensure that population
numbers were not allocated to areas where people do not live in Hamilton
(for example, to parks or large institutional buildings, etc.). To do
so, we retrieved shapefiles for various geographic features {[}see Table
\ref{tab:data-features}{]} from Open Hamilton. Next, we removed these
features from the PBSP core service area and used pycnophylactic
interpolation to disaggregate and reallocate population within the
remaining area {[}see Figure \ref{fig:interpolated-population}{]}.

@@

\caption{Interpolated population in the Hamilton Bike Share's core service area (outlined in black) and within 30 minutes of walking to the core service area. Each population cell is 50-by-50 m in size. The downtown area is outlined in dark green.}\label{fig:interpolated-population}
\end{figure}

@@

To calculate walking times from the centroid of the micro population
cells to docking stations, we extracted OpenStreetMap data for Hamilton
Bike Share's service area from
\href{https://download.bbbike.org/osm/bbbike/}{BBBike}, an online cycle
route planner that interfaces with OpenStreetMap. OpenStreetMap data
provides the networks for calculating walking times from each population
cell to nearby bicycle share stations, using a maximum walking distance
of 10 km and walking time of 30 minutes as thresholds. A travel time
matrix was created with the origins as the coordinates of the population
cells and the destinations as the coordinates of the bicycle share
stations within the maximum threshold. This process provides a more
realistic measure of the friction of reaching stations by taking
infrastructure into account network travel times, rather than using the
Euclidean distance from population cell to station. Routing and travel
time calculations were completed using the \texttt{R} package
\texttt{r5r}, used for rapid realistic routing operations (Pereira et
al., 2021).

@@

All data for this research were accessed from publicly available Census
of Canada sources, from OpenStreetMaps, and from Open Hamilton\footnote{https://open.hamilton.ca/},
a public online repository of data curated by the City of Hamilton.

@@

Consensus regarding the distance that individuals are willing to walk to
access a bike share station is lacking, but the literature on walking
behaviour can provide some guidelines to determine the thresholds for
our sensitivity analysis. Previous studies have found that living within
250 m (Fuller et al., 2011) and 300 m (Kabra et al., 2020) of a docking
station is correlated with bicycle share use, while other research has
found that walking trips are less than 600 m and rarely more than 1200 m
(Millward et al., 2013) or a median distance of 650 m (Larsen et al.,
2010). Hamilton Bike Share will often depict a map at some docking
stations to show the locations of the other nearest stations within a
five minute walk, which suggests that this is an average distance that
people are expected to walk. National Association of City Transportation
Officials (NACTO) has a similar normative guide (City Transportation
Officials, 2015).

In the present case, we found that accessibility calculated using the
balanced FCA approach increases with a threshold between two and four
minutes, but is then maximized at five minutes. Accessibility decreases
substantially after eight minutes, which is intuitive given that demand
on a limited supply increases as more people can reach each station.

For this reason, we experimented with various thresholds by conducting a
sensitivity analysis to calculate accessibility at different walking
times from population cell to bike share stations: three minutes, five
minutes, ten minutes, and fifteen minutes. We categorized these
thresholds as minimum, average, maximum, and extreme, respectively. At
each threshold, we compared accessibility between the current system and
the original system to examine the contribution of the twelve equity
stations. When considering the results reported below, it is important
to remember that accessibility is technically a form of smoothing
(O'Kelly and Horner, 2003, pp. 7--8): smaller thresholds produce less
smoothing (which can result in ``spiky'' accessibility landscapes),
while larger thresholds produce more smoothing and fewer spikes.

@@

With a walking distance of fifteen minutes, we found that there are 2.44
bicycle racks per person system-wide in the original system
configuration (i.e., without equity stations). With the addition of
equity stations, there are now 2.55 bicycles per person. Figure
\ref{fig:figure-9} presents a comparison of accessibility between the
systems. Users near the university and the neighborhoods north of the
downtown area (the latter is outlined in green) have the highest
accessibility, followed by those who live in the city's downtown area.
Accessibility in the east end, where equity stations were implemented,
of the core service area remains low.

@@

To highlight the benefit of using the BFCA approach, we conducted a
comparative analysis with the 2SFCA method. The latter method does not
allocate population and level of service proportionally, which means
that there are often substantial amounts of multiple counting each.
Figure \ref{fig:figure-10} shows the level of inflation and deflation at
five minutes that results when the balanced and the conventional
approaches are compared. The inflation and deflation is defined as the
ratio of the level of service and accessibility calculated using the
conventional approach to the balanced approach. This effect has been
documented in other studies (Chen et al., 2021; Paez et al., 2019).
Figure \ref{fig:figure-10} shows that in some locations the level of
service is deflated by up to 80\%. This is evident for the equity
stations in the east end of the core service area. Noticeably, the level
of service is least deflated in the downtown area, likely due to the
density of stations. Figure \ref{fig:figure-11} shows that accessibility
is inflated by as much as 50\% in many areas across the city.

@@

Next, we wish to examine accessibility differences by income level. To
this end, we took our micro-zones and imputed the median household
income of their corresponding DA. Use of bivariate maps helps us to
visualize the level of accessibility by level of income. Median total
household income statistics were drawn from the 2016 Canadian Census.

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Examining equity in accessibility to bike share: a balanced floating catchment area approach},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Examining equity in accessibility to bike share: a balanced floating
catchment area approach}
    \author[McMaster University]{Elise Desjardins\corref{1}}
   \ead{desjae@mcmaster.ca} 
    \author[University of Toronto Scarborough]{Christopher D. Higgins}
   \ead{cd.higgins@utoronto.ca} 
    \author[McMaster University]{Antonio Paez}
   \ead{paezha@mcmaster.ca} 
      \address[McMaster University]{School of Earth, Environment \& Society, McMaster University, 1280 Main
Street West, Hamilton, ON L8S4L8}
    \address[University of Toronto Scarborough]{Department of Geography \& Planning, University of Toronto Scarborough,
1265 Military Trail, Toronto, ON M1C1A4}
      \cortext[1]{Corresponding Author}

@@

Many PBSPs in North America now offer specific programs to address
equity that primarily focus on removing cost barriers and increasing
access for groups that are under-represented among existing users
(McNeil et al., 2019). The locations of docking stations is a common
consideration for increasing equity (Howland et al., 2017). Hamilton
Bike Share (HBS), located in Hamilton, Ontario, was the only Canadian
PBSP included in a North American scan of bike share equity programs
(see McNeil et al., 2019). HBS was launched in 2015 and currently has
over 900 operational bicycles and 130 docking stations. An equity
program, Everyone Rides Initiative, was implemented in 2018 which
expanded the program by introducing twelve ``equity'' stations to more
disadvantaged neighbourhoods in the core service area.

This paper examines accessibility to Hamilton Bike Share using a
balanced floating catchment area (BFCA) approach. We also conduct a
comparative analysis to the conventional two-step floating catchment
area (2SFCA) approach to highlight the benefits of this method which, to
our knowledge, has not yet been used in the cycling literature. The
paper also assesses the contribution of the equity stations to reducing
inequities in accessibility for different groups according to median
total household income, and provides policy recommendations to further
improve equity.

@@

\hypertarget{literature-review}{%
\section{2. Literature Review}\label{literature-review}}

\hypertarget{public-bicycle-share-programs}{%
\subsection{2.1. Public Bicycle Share
Programs}\label{public-bicycle-share-programs}}

@@

\hypertarget{equity-of-pbsps}{%
\subsection{2.2. Equity of PBSPs}\label{equity-of-pbsps}}

@@

On the whole, existing studies highlight the need for PBSPs to be more
accessible for diverse populations in order to increase use beyond the
``typical'' users. This has been the focus of recent research (see,
among others, Auchincloss et al., 2020; Hull Grasso et al., 2020;
MacArthur et al., 2020). Offering more people the option of using
sustainable and active transportation, particularly those who have lower
socioeconomic status and might benefit the most, is a worthy policy goal
for cities with PBSPs. However, exploring transportation equity by
investigating where bike share stations are located, often using
neighborhoods or census tracts as the geographical unit of analysis, can
ignore or miss the benefits that may be derived from adjacent zones.
Meaning that, stations may be lacking in certain neighborhoods but there
may be stations accessible within a reasonable walking time. This is
where geographical accessibility becomes an important consideration.

@@

The location and size of docking stations are two factors that are
relevant to accessibility. Since the time or distance needed to reach a
docking station decreases the potential of accessing the program, the
location matters. Kabra et al.~(2020) found that the majority of bike
share usage in Paris, France comes from areas within 300 m of stations,
which amounts to 2-4 minutes walking by an adult who does not have a
disability. Similar to other public amenities affected by crowding, the
utility of docking stations is also limited by the \emph{maximum} number
of bicycles that they can hold (e.g., their size). Accessibility
analyses for PBSPs constitute a positive and evaluation-based approach
that also has the potential to inform equity efforts. For instance, Wang
and Lindsey (2019b) investigated whether new or relocated bike share
stations increased accessibility and use, which offered important
insights to improve the performance of the program.

Several approaches have been commonly used for measuring place-based
accessibility, including cumulative opportunities, gravity, and
utility-based measures (Handy and Niemeier, 1997). The gravity-based
approach involves weighting destination opportunities, such as the
quantity of bike share stations, by the time required to reach them from
an origin using an impedance function (Handy and Niemeier, 1997; Kwan,
1998). While such measures are suitable for capturing the potential for
reaching destinations from a given location, they do not take demand or
congestion effects into account which is an important consideration when
calculating accessibility for amenities such as bike share stations.

@@

When measuring accessibility, researchers have also taken different
approaches when it comes to the aggregation of data, either by using the
individual or household as the smallest unit of analysis or larger
spatial zones. Previous research on bike share equity has typically used
a meso- or macro-level approach with aggregated data from entire
neighborhoods or census tracts (Babagoli et al., 2019; Mooney et al.,
2019; Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), although there
are recent exceptions (Chen et al., 2019; Chen and Li, 2021). This is
also true for studies examining correlates of bike share demand (J. Wang
and Lindsey, 2019b). Handy and Niemeier (1997) note that using
disaggregated data in accessibility analyses provides a more accurate
estimate for individuals. Chen et al.~(2019, p. 530) are in favour of
using disaggregated data, which they did in their recent analysis of
Tampa's PBSP, because they note ``the use of aggregated data might
hinder our understanding of the equity impacts since individual
disparities are absorbed after aggregation''.

@@

Our analysis builds upon a previous study (Hosford and Winters, 2018),
which found that areas in Hamilton with less advantage are better served
by the city's PBSP compared to other Canadian cities (i.e., Toronto,
Vancouver, Montreal, and Ottawa-Gatineau) where areas that are less
deprived have greater access. Hosford and Winters (2018, p. 47)
acknowledge that ``Hamilton stands out in that the lower income
neighborhoods are located near the city center and wealthier
neighborhoods are in the surrounding suburban areas''. Therefore, the
core service area for the PBSP in Hamilton by default covers more of the
disadvantaged areas in the city. However, there is also a great deal of
variation in income in the core service area because of the local
university and increasing gentrification. Hosford and Winters (2018)
took a macro-level approach in their analysis by using dissemination
areas \emph{across} the city as the unit of analysis. They did not focus
specifically on the core service area and did not differentiate between
conventional and equity stations.

Therefore, this paper also contributes to the cycling literature by
expanding the analysis of Hosford and Winters (2018) to assess equity in
accessibility to HBS and explore the contribution of the equity stations
added to Hamilton's PBSP.

@@

The case study of this paper is the city of Hamilton, Ontario, Canada.
Before June 2020, the program was known as Social Bicycles or SoBi
Hamilton, but is now called Hamilton Bike Share (HBS). The core service
area spans 40 sq.km of the city, although it was planned to be 20 sq.km
(Hamilton, 2015a), and roughly 138,000 people are within 30 minutes
walking of a station {[}see Figure
\ref{fig:hamilton-and-sobi-service-area}{]}. This represents roughly one
fifth of the total population of the Hamilton Census Metropolitan Area
according to the 2016 Canadian Census. The City of Hamilton undertook a
large public engagement campaign to validate the locations of stations
that had been selected and to crowdsource potential locations for
additional locations (Hamilton, 2014). Most of the potential locations
that were suggested were in the east end of the core service area that
lack stations or in neighborhoods not serviced by the PBSP. The program
was enthusiastically welcomed in the city in 2015 - within three weeks
of launching, 10,000 trips had been made (Hamilton, 2015b), however
inadequate coverage given the size of the program's service area was
identified as a problem early on, and transportation planners noted that
the small size (i.e., supply of bicycle racks) and low quantity of
stations would lead to challenges in balancing the system (Hamilton,
2015a).

@@

Hamilton Bike Share Inc., the non-profit operator, launched an equity
program, Everyone Rides Initiative (ERI), in 2018 with the objective of
reducing barriers that may prevent individuals from accessing the
program. Additional bicycles and twelve ``equity'' stations were added
to the core service area in more disadvantaged neighbourhoods. The
equity program also offers subsidized memberships to individuals who
identify as low income, and complements this service with cycle skills
education. A comparable program can be found in Philadelphia (see Caspi
and Noland, 2019).

@@

\caption{The spatial distribution of bike share docking stations in Hamilton, Ontario. Conventional stations are in purple and equity (ERI) stations are in orange. The service area of Hamilton Bike Share is outlined in blue and the city's downtown core is outlined in dark green. Hamilton Census Metropolitan Area is shown in grey.}\label{fig:sobi-stations-in-hamilton}
\end{figure}

@@

Since walking trips to a docking station likely happen at a level lower
than even the smallest census geography, this requires a more
disaggregated approach than the use of census geographies. For this
reason, we implemented our analysis parting from micro population zones
to better reflect the friction of walking to a docking station, which is
an important component of a bike share trip (Chen et al., 2019).

To obtain population at sub-census geography levels, we used
pycnophylactic interpolation (Tobler, 1979). We obtained population data
from the 2016 Census of Canada for dissemination areas (DA), which is
the smallest publicly available census geography in Canada. These zonal
values of the population were interpolated to smaller polygons that are
50-by-50 m in size. Pycnophylactic interpolation involves smoothing out
the population from each DA while preserving total volume {[}see Figure
\ref{fig:da-population}{]}. When interpolating the population at this
high level of resolution, it is important to ensure that population
numbers were not allocated to areas where people do not live in Hamilton
(for example, to parks or large institutional buildings, etc.). To do
so, we retrieved shapefiles for various geographic features from Open
Hamilton. Next, we removed these features from the PBSP core service
area and used pycnophylactic interpolation to disaggregate and
reallocate population within the remaining area {[}see Figure
\ref{fig:interpolated-population}{]}.

@@

\caption{Interpolated population in Hamilton Bike Share's core service area (outlined in blue) and within 30 minutes of walking to the core service area. Each population cell is 50-by-50 m in size. The downtown area is outlined in dark green.}\label{fig:interpolated-population}
\end{figure}

@@

To calculate walking times from the centroid of the micro population
cells to docking stations, we extracted OpenStreetMap data for HBS's
service area from
\href{https://download.bbbike.org/osm/bbbike/}{BBBike}, an online cycle
route planner that interfaces with OpenStreetMap. OpenStreetMap data
provides the networks for calculating walking times from each population
cell to nearby docking stations, using a maximum walking distance of 10
km and walking time of 30 minutes as thresholds. A travel time matrix
was created with the origins as the coordinates of the population cells
and the destinations as the coordinates of the docking stations within
the maximum threshold. This process provides a more realistic measure of
the friction of reaching stations by taking infrastructure into account
network travel times, rather than using the Euclidean distance from
population cell to station. Routing and travel time calculations were
completed using the \texttt{R} package \texttt{r5r}, used for rapid
realistic routing operations (Pereira et al., 2021).

@@

All data for this research were accessed from publicly available Census
of Canada sources, from OpenStreetMaps, and from Open Hamilton\footnote{https://open.hamilton.ca/},
a public online repository of data curated by the City of Hamilton.
Median total household income statistics were drawn from the 2016
Canadian Census.

@@

Consensus regarding the distance that individuals are willing to walk to
access a docking station is lacking, but the literature on walking
behaviour provided some guidelines to determine the thresholds for our
sensitivity analysis. Previous studies have found that living within 250
m (Fuller et al., 2011) and 300 m (Kabra et al., 2020) of a docking
station is correlated with bike share use, while other research has
found that walking trips are less than 600 m and rarely more than 1200 m
(Millward et al., 2013) or a median distance of 650 m (Larsen et al.,
2010). HBS will often depict a map at some docking stations to show the
locations of the other nearest stations within a five minute walk, which
suggests that this is an average distance that people are expected to
walk. The National Association of City Transportation Officials (NACTO)
has a similar normative guide (City Transportation Officials, 2015).

In the present case, we found that accessibility calculated using the
BFCA method increased with a threshold between two and four minutes, but
was then maximized at five minutes. Accessibility decreased
substantially after eight minutes, which is intuitive given that demand
on a limited supply increases as more people can reach each station.

We experimented with various thresholds by conducting a sensitivity
analysis to calculate accessibility at different walking times from
population cell to docking stations: three minutes, five minutes, ten
minutes, and fifteen minutes. We categorized these thresholds as
minimum, average, maximum, and extreme, respectively. At each threshold,
we compared accessibility between the current system and the original
system to examine the contribution of the twelve equity stations. When
considering the results reported below, it is important to remember that
accessibility is technically a form of smoothing (O'Kelly and Horner,
2003, pp. 7--8): smaller thresholds produce less smoothing (which can
result in ``spiky'' accessibility landscapes), while larger thresholds
produce more smoothing and fewer spikes.

@@

With a walking distance of fifteen minutes, we found that there are 2.44
bicycle racks per person system-wide in the original system
configuration (i.e., without equity stations). With the addition of
equity stations, there are now 2.55 bicycles per person. Figure
\ref{fig:figure-9} presents a comparison of accessibility between the
systems. Users near the university and the neighborhoods north of the
downtown area (the latter is outlined in green) have the highest
accessibility, followed by those who live in the city's downtown area.
Accessibility in the east end, where equity stations were implemented,
of the core service area remains lower than other areas.

@@

To highlight the benefit of using the BFCA approach, we conducted a
comparative analysis with the 2SFCA method. The latter method does not
allocate population and level of service proportionally, which means
that there are often substantial amounts of multiple counting each.
Figure \ref{fig:figure-10} shows the level of inflation and deflation at
five minutes that results when the balanced and the conventional FCA
methods are compared. The inflation and deflation is defined as the
ratio of the level of service and accessibility calculated using the
conventional approach to the balanced approach. This effect has been
documented in other studies (Chen et al., 2021; Paez et al., 2019).
Figure \ref{fig:figure-10} shows that in some locations the level of
service is deflated by up to 80\%. This is evident for the equity
stations in the east end of the core service area. Noticeably, the level
of service is least deflated in the downtown area, likely due to the
density of docking stations. Figure \ref{fig:figure-11} shows that
accessibility is inflated by as much as 50\% in many areas across the
city.

@@

To examine whether accessibility to HBS was increased for groups with
lower income, we estimated what level of accessibility accrues to how
many people in different income strata. To this end, we took our micro
population zones and aggregated the accessibility and population by the
median total household income as imputed from the dissemination areas. A
unique property of the BFCA method, which does not hold for the 2SFCA
approach because of the inflation/deflation issues discussed above, is
that accessibility and population data can be re-aggregated using income
as an aggregator criterion, while preserving the total population and
the level of service. This avoids demand and supply inflation, and also
enables us to present findings in a way that is more intuitive to
interpret. Figures \ref{fig:figure-bi-map-threshold-3},
\ref{fig:figure-bi-map-threshold-5},
\ref{fig:figure-bi-map-threshold-10}, and
\ref{fig:figure-bi-map-threshold-15} depict bivariate choropleth maps
that combine the spatial distribution of accessibility and median total
household income, using tertiles for the coloring scheme.

\begin{figure}
\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-3-1} \caption{\label{fig-bivariate-map-threshold-3}Bivariate map of accessibility and income at the minimum threshold of three minutes with equity stations (top panel) and without equity stations (bottom panel).}\label{fig:figure-bi-map-threshold-3}
\end{figure}

\begin{figure}
\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-5-1} \caption{\label{fig-bivariate-map-threshold-5}Bivariate map of accessibility and income at the average threshold of five minutes with equity stations (top panel) and without equity stations (bottom panel).}\label{fig:figure-bi-map-threshold-5}
\end{figure}
